My wonderful paper

### 2024-12-06 _manuscript/_tex/index.tex
smooth abstract till Section 3.1

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
Unexpected results in data analysis can prompt researchers to examine
data quality and analysis steps. While some researchers can typically
diagnose issues effectively with a few checks, others may struggle to
identify appropriate checks to diagnose the problem. {[}an example of
check{]} These checks are often informal and difficult to trace and
discuss in publications, resulting in others questioning the
trustworthiness of the analysis. To address this, we formalize the
informal checks into an \emph{analysis plan} that encompasses the
analysis steps and (the unit tests): one test for whether the result
meets expectations and multiple tests for checking the analysis. We then
present a procedure to assess the quality of these unit tests based on
their accuracy and redundancy on simulated versions of the original
data. The accuracy is assessed using binary classification metrics,
\emph{i.e.}, precision and recall, while redundancy is calculated using
mutual information. This procedure can be used to conduct a sensitivity
analysis, compare different analysis plans, and to identify the optimal
cutoff point for the unit tests.
\end{abstract}

@@

One might gain insight into analysts' thought process by speaking with
them directly or watching screencast videos they produce, such as,
TidyTuesday screencast videos. However, direct conversation are not
scalable and may not be feasible while creating educational videos
requires significant effort from the researchers. Ideally, there could
be a way to make these expectations explicit and accessible to others.
Even better, if the encoding is machine-readable, we could analyze these
expectations and learn from the analysis -- whether the checks also
apply to other researchers analyzing new data in the same context,
whether they reflect commo common practices in the field, or whether
they are sepcific to the data or analysis at hand.

In this paper, we frame these expectations as checks within an
\emph{analysis plan}, which allows us to examine these implicit
assumptions made about the data during analysis and to compare different
checks for diagnosing unexpected outcomes. We then introduce a procedure
to evaluate the quality of these checks, either individually or in
combination through logic regression, based on their accuracy and
redundancy across simulation of the original data. Accuracy is assessed
using binary classification metrics, precision and recall, while
redundancy is measured using mutual information. The proposed workflow
offers a numerical guarantee, based on simulation, that the analysis
will produce the expected results, assuming the data generating
mechanism holds. {[}something about trustworthy data science{]}

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concept of diagnosing
unexpected outcomes and data quality checks. Section~\ref{sec-plan}
introduces the concept of analysis plan, illustrated with a toy example
on fitness step count. Section~\ref{sec-method} describes the procedure
for selecting optimal check combinations to predict unexpected outcomes.
Section~\ref{sec-pm10-mortality} applies this selection procedure to a
linear regression example examining the effect of PM10 on mortality.
Section~\ref{sec-discussion} discusses a few key considerations and
Section~\ref{sec-conclusion} concludes the paper.

@@

They shape how we interpret the results and assess whether they are
consistent with existing knowledge or indicate the need for updates.
\citep{grolemund_cognitive_2014}

\citep{peng_diagnosing_2021} describes three pedagogical exercises of
introducing diagnosing unexpected outcome into a classroom setting.

TODO: what if the expectation is ``wrong''

@@

Expectations represent our beliefs about certain aspects of the
analysis, independent of the analysis itself. When outcomes deviate from
these expectations, analysts often revisit the analysis cycle to
identify potential issues, refine methods, or revise assumptions.
Experienced analysts can typically identify issues quickly and correct
them on the spot, but they often do so without discussing the underlying
reasoning, making it harder for less experienced researchers to learn
and master these skills. Here, we introduce the concept of
\emph{analysis validation checks}, which frames these expectations or
assumptions as explicit checks that returns a TRUE or FALSE output given
the data analyzed at hand. Inspired by \emph{data validation checks}
\citep{validate}, where ensure datasets meet expected format and
quality, analysis validation checks reverse the approach: they validate
the assumptions about the data necessary for the analysis to produce the
reported results. This allows the concept to encompass a broad range of
checks, such as data quality (i.e.~missing data, how the data is
structured), data distribution and outliers, bivariate and multivariate
relationships between variables, and contextual information. These
externalized checks provide insights into analysts' thoughts process
process and offer several benefits:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  help students and junior researchers develop skills to diagnose
  unexpected analysis results,
\item
  serve as clear checkpoints to support independent replication or
  application of methods to new data, and
\item
  align assumptions among researchers from different domain backgrounds.
\item
  improve analysis transparency, reproducibility, and trustworthiness.
\end{enumerate}

@@

\section{Applications}\label{sec-examples}

A regression example is produced to illustrate the test selection
process for analyzing the effect of PM10 on mortality. The example
demonstrates how the test process can be used to select cutoff values in
the unit tests and how the test can be used to diagnose an unexpected
PM10 coefficient from the generalized linear model.

\subsection{Effect of PM10 on mortality}\label{sec-pm10-mortality}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While experienced
researchers can typically diagnose problems by checking a few key
assumptions, others may struggle to identify the root causes. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
formalize these informal assumptions by framing them as binary
\emph{analysis validation check}. We then introduce a procedure to
quantify how violations of these checks may lead to unexpected results
in the analysis. The procedure relies on simulations of the original
data and evaluates boht accuracy and redundancy. Accuracy is calculated
through binary classification metric, while redundancy is measured using
mutual information. We demonstrate this approach with a toy example on
the fitness step count data and a generalized linear model exmaple
examining the effect of PM10 on mortality.
\end{abstract}

@@

One might gain insight into analysts' thought process by speaking with
them directly or watching screencast videos they produce, such as,
TidyTuesday screencast videos. However, direct conversation are not
scalable and may not be feasible while creating educational videos
requires significant effort from the researchers. Ideally, there could
be a way to make these expectations explicit and accessible to others.
Even better, if the encoding is machine-readable, we could analyze these
expectations and learn from the analysis -- whether the checks also
apply to other researchers analyzing new data in the same context,
whether they reflect commo common practices in the field, or whether
they are specific to the data or analysis at hand.

In this paper, we conceptualize these internal expectations and
assumptions as \emph{analysis validation checks}, which allows us to
examine these assumptions made during analysis to diagnose unexpected
outcomes. We then introduce a procedure that provides a quantitative
emasure on how violations in a tree of checks, derived from individual
checks, will lead to an unexpected result. The procedure, based on
simulations of the original data, calculates accuracy and redundancy.
Accuracy is determined using binary classification metrics, precision
and recall, from a logic regression fit while redundancy is measured
using mutual information. The proposed workflow offers a numerical
guarantee that the analysis will produce the expected results, assuming
the data generating mechanism holds. {[}something about trustworthy data
science{]}

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concept of diagnosing
unexpected outcomes and data quality checks. Section~\ref{sec-plan}
introduces the concept of analysis validation check, illustrated with a
toy example on fitness step count. Section~\ref{sec-method} describes
the procedure that quantify how checks combined in logical operators can
predict unexpected outcomes. Section~\ref{sec-pm10-mortality} applies
this procedure to examine the effect of PM10 on mortality in public
health in a generalized linear model. Section~\ref{sec-discussion}
discusses a few key considerations and Section~\ref{sec-conclusion}
concludes the paper.

@@

\begin{itemize}
\item
  They shape how we interpret the results and assess whether they are
  consistent with existing knowledge or indicate the need for updates.
  \citep{grolemund_cognitive_2014}
\item
  \citep{peng_diagnosing_2021} describes three pedagogical exercises of
  introducing diagnosing unexpected outcome into a classroom setting.
\item
  what if the expectation is ``wrong''
\end{itemize}

@@

Expectations represent our beliefs about certain aspects of the
analysis, independent of the analysis itself. When outcomes deviate from
these expectations, analysts often revisit the analysis cycle to
identify potential issues, refine methods, or revise assumptions.
Experienced analysts can typically identify issues quickly and correct
them on the spot, but they often do so without discussing the underlying
reasoning, making it harder for less experienced researchers to learn
and master these skills. Here, we introduce the concept of
\emph{analysis validation checks}, which frames these expectations or
assumptions as explicit checks that returns a TRUE or FALSE output given
the data analyzed at hand. Inspired by data validation checks
\citep{validate}, where ensure datasets meet expected format and
quality, analysis validation checks reverse the approach: they validate
the assumptions about the data necessary for the analysis to produce the
reported results. This allows the concept to encompass a broad range of
checks, such as data quality (i.e.~missing data, how the data is
structured), data distribution and outliers, bivariate and multivariate
relationships between variables, and contextual information. These
externalized checks provide insights into analysts' thoughts process
process and offer several benefits:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  help students and junior researchers develop skills to diagnose
  unexpected analysis results,
\item
  serve as clear checkpoints to support replication or application of
  methods to (new) data,
\item
  align assumptions among researchers from different domain backgrounds,
  and
\item
  improve analysis transparency, reproducibility, and trustworthiness.
\end{enumerate}

@@

\section{Application}\label{sec-pm10-mortality}

A regression example is produced to illustrate the test selection
process for analyzing the effect of PM10 on mortality. The example
demonstrates how the procedure presented in Section~\ref{sec-method} can
be used to select cutoff values in the checks to diagnose an unexpected
PM10 coefficient from the generalized linear model.

### 2024-12-06 index.qmd
smooth abstract till Section 3.1

---
title: "My wonderful paper"
abstract: "Unexpected results in data analysis can prompt researchers to examine data quality and analysis steps. While some researchers can typically diagnose issues effectively with a few checks, others may struggle to identify appropriate checks to diagnose the problem. [an example of check] These checks are often informal and difficult to trace and discuss in publications, resulting in others questioning the trustworthiness of the analysis. To address this, we formalize the informal checks into an *analysis plan* that encompasses the analysis steps and (the unit tests): one test for whether the result meets expectations and multiple tests for checking the analysis. We then present a procedure to assess the quality of these unit tests based on their accuracy and redundancy on simulated versions of the original data. The accuracy is assessed using binary classification metrics, *i.e.*, precision and recall, while redundancy is calculated using mutual information. This procedure can be used to conduct a sensitivity analysis, compare different analysis plans, and to identify the optimal cutoff point for the unit tests."
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
    attributes:
        corresponding: true
  - name: Roger D. Peng
    email: roger.peng@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
cite-method: natbib
bibliography: references.bib
format:
 tandf-pdf:
   keep-tex: true
   fontsize: 12pt
   linestretch: 2
   notebook-links: false
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

One might gain insight into analysts' thought process by speaking with them directly or watching screencast videos they produce, such as, TidyTuesday screencast videos. However, direct conversation are not scalable and may not be feasible while creating educational videos requires significant effort from the researchers. Ideally, there could be a way to make these expectations explicit and accessible to others. Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are sepcific to the data or analysis at hand.

In this paper, we frame these expectations as checks within an *analysis plan*, which allows us to examine these implicit assumptions made about the data during analysis and to compare different checks for diagnosing unexpected outcomes. We then introduce a procedure to evaluate the quality of these checks, either individually or in combination through logic regression, based on their accuracy and redundancy across simulation of the original data. Accuracy is assessed using binary classification metrics, precision and recall, while redundancy is measured using mutual information. The proposed workflow offers a numerical guarantee, based on simulation, that the analysis will produce the expected results, assuming the data generating mechanism holds. [something about trustworthy data science]

The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks. @sec-plan introduces the concept of analysis plan, illustrated with a toy example on fitness step count. @sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes. @sec-pm10-mortality applies this selection procedure to a linear regression example examining the effect of PM10 on mortality. @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.

@@

They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]

[@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.

TODO: what if the expectation is "wrong"

@@

<!-- We may anticipate certain interval on the main result of the analysis based on prior knowledge. For example, researchers in public health may expect an average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient] [TODO: check if this is correct].  -->

Expectations represent our beliefs about certain aspects of the analysis, independent of the analysis itself. When outcomes deviate from these expectations, analysts often revisit the analysis cycle to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills. Here, we introduce the concept of *analysis validation checks*, which frames these expectations or assumptions as explicit checks that returns a TRUE or FALSE output given the data analyzed at hand. Inspired by *data validation checks* [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results. This allows the concept to encompass a broad range of checks, such as data quality (i.e. missing data, how the data is structured), data distribution and outliers, bivariate and multivariate relationships between variables, and contextual information. These externalized checks provide insights into analysts' thoughts process process and offer several benefits:

  1) help students and junior researchers develop skills to diagnose unexpected analysis results, 
  2) serve as clear checkpoints to support independent replication or application of methods to new data, and 
  3) align assumptions among researchers from different domain backgrounds.
  4) improve analysis transparency, reproducibility, and trustworthiness.

@@

# Applications {#sec-examples}

A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality. The example demonstrates how the test process can be used to select cutoff values in the unit tests and how the test can be used to diagnose an unexpected PM10 coefficient from the generalized linear model.

## Effect of PM10 on mortality {#sec-pm10-mortality}

>>>>

---
title: "My wonderful paper"
abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues. While experienced researchers can typically diagnose problems by checking a few key assumptions, others may struggle to identify the root causes. These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications. In this paper, we formalize these informal assumptions by framing them as binary *analysis validation check*. We then introduce a procedure to quantify how violations of these checks may lead to unexpected results in the analysis. The procedure relies on simulations of the original data and evaluates boht accuracy and redundancy. Accuracy is calculated through binary classification metric, while redundancy is measured using mutual information. We demonstrate this approach with a toy example on the fitness step count data and a generalized linear model exmaple examining the effect of PM10 on mortality. "
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
    attributes:
        corresponding: true
  - name: Roger D. Peng
    email: roger.peng@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
cite-method: natbib
bibliography: references.bib
format:
 tandf-pdf:
   keep-tex: true
   fontsize: 12pt
   linestretch: 2
   notebook-links: false
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

One might gain insight into analysts' thought process by speaking with them directly or watching screencast videos they produce, such as, TidyTuesday screencast videos. However, direct conversation are not scalable and may not be feasible while creating educational videos requires significant effort from the researchers. Ideally, there could be a way to make these expectations explicit and accessible to others. Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are specific to the data or analysis at hand.

In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine these assumptions made during analysis to diagnose unexpected outcomes. We then introduce a procedure that provides a quantitative emasure on how violations in a tree of checks, derived from individual checks, will lead to an unexpected result. The procedure, based on simulations of the original data, calculates accuracy and redundancy. Accuracy is determined using binary classification metrics, precision and recall, from a logic regression fit while redundancy is measured using mutual information. The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the data generating mechanism holds. [something about trustworthy data science]

The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks. @sec-plan introduces the concept of analysis validation check, illustrated with a toy example on fitness step count. @sec-method describes the procedure that quantify how checks combined in logical operators can predict unexpected outcomes. @sec-pm10-mortality applies this procedure to examine the effect of PM10 on mortality in public health in a generalized linear model. @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.

@@

* They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]

* [@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.

* what if the expectation is "wrong"

@@

Expectations represent our beliefs about certain aspects of the analysis, independent of the analysis itself. When outcomes deviate from these expectations, analysts often revisit the analysis cycle to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills. Here, we introduce the concept of *analysis validation checks*, which frames these expectations or assumptions as explicit checks that returns a TRUE or FALSE output given the data analyzed at hand. Inspired by data validation checks [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results. This allows the concept to encompass a broad range of checks, such as data quality (i.e. missing data, how the data is structured), data distribution and outliers, bivariate and multivariate relationships between variables, and contextual information. These externalized checks provide insights into analysts' thoughts process process and offer several benefits:

  1) help students and junior researchers develop skills to diagnose unexpected analysis results, 
  2) serve as clear checkpoints to support replication or application of methods to (new) data,
  3) align assumptions among researchers from different domain backgrounds, and 
  4) improve analysis transparency, reproducibility, and trustworthiness.

@@

# Application {#sec-pm10-mortality}

A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality. The example demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
